---
title: 'Need help managing your references? "El Zotero" to the rescue!'
description: |
  In the Lands in Between bibliographies, select Zotero as your hero.
author:
  - name: Juan C. Sanchez-Arias
    url: https://www.juansanar.com/
date: 2022-07-04
output:
  distill::distill_article:
    self_contained: false
draft: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## It's 2022! Use a reference manager ffs

The tradition of passing information to one generation another one has allowed us to use knowledge to our advantage (TODO REF). However, successfully achieving this requires the ability to track down the origins of such knowledge and related observations, opinions, etc. by adding small notes about them within our writing (TODO REF) as references or footnotes. In the past, this was done manually of course (and you'll be surprised how many people still follow this ice age practice) and with the development of technology (e.g., personal computers, text processing software, etc), this labour intensive task can be somewhat automated (you still need to gather the information and make sense of it of course, though AI is taking major leaps in this regard).

Another challenge that accumulating knowledge brings is managing the sheer amount of content. I wrote a small tidbit about the advantages of using a personal knowledge management tool. In this post, I will focus on managing references, specifically, I will try to convince you should use a reference manager for:

- Organizing your information sources (e.g., articles, books chapters, dissertations, newspapers, etc).
- Inserting and formatting citations into your text editor of preference (e.g., MS Word, LibreOffice, Google Docs, Zettlr, VS Code).
- Creating and formatting a bibliography.

**So how do we deal with all this information?**
Specialized software such as reference managers are fundamental aids when citing knowledge in our writing, especially within the ivory walls of academia, where we ("we" as the academic collective through history) have created rules and style formats for this purpose, to the point that almost every single scholarly journal produces a format that best suites their copyediting and styling.

## How to choose your reference manager?

When it comes to choosing any piece of software, I like to use the following criteria to guide my decision:

- Is it FOSS (**F**ree **O**pen **S**ource **S**oftware)?
  - I gravitate towards FOSS because it means I can see how the application works, but also I can contribute to it and take advantage of others contributions (which tend to come in amazing plugins and customization options). Also, who doesn't like free stuff?

- Does it have a community backing up or supporting the application?
  - This is really important because it makes troubleshooting bugs a breeze and not an insurmountable headache.

- Does it support multiple operating systems and multiple document writing software? In other words: does it work on macOS, MS Windows, and Linux? Does it work with MS Word, Google Docs, LibreOffice, and others (e.g., VS Code, Zettlr)?
  - As someone that uses all of these 3 operating systems (macOS for work and MS Windows and Linux for personal use) and multiple writing applications (primairly MS Word, Google Docs, and VS Code), this is a critical point.

In my case, Zotero check all the boxes and beyond, and it has been and continues to be my primary recommendation in the internet (e.g., [here](https://www.reddit.com/r/bioinformatics/comments/vcs95u/comment/icgmr9h/?utm_source=share&utm_medium=web2x&context=3), [here](https://www.reddit.com/r/academia/comments/txjo0p/comment/i3m4pod/?utm_source=share&utm_medium=web2x&context=3), [here](https://twitter.com/juan_sanar/status/1082454881531842560?s=20&t=wd1vwP0KEX2yXQx_Dr-5Uw), [here](https://www.reddit.com/r/academia/comments/mao33p/comment/gru8g0r/?utm_source=share&utm_medium=web2x&context=3), [here](https://www.reddit.com/r/PhD/comments/js4z40/comment/gbyjs61/?utm_source=share&utm_medium=web2x&context=3), and so on).

You get my point, I love Zotero, advocate for Zotero, and no I am NOT sponsored by Zotero but I have immensley benefited from its flexibility and cross-platform compatbility. However, in the interest of experimentation, I'll briefly summarize the major names in the reference manager game (at least from what I've learned from reading and talking to primarily other folks in academia - there's another way to manage your reference if you're also a fan of plain text like me, but that's for another day-)

```{r panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

::::: {.panelset}

::: {.panel}

### Zotero {.panel-name}

asdf

:::

::: {.panel}

### Mendeley {.panel-name}

asdf

:::

::: {.panel}

### EndNote {.panel-name}

asdf

:::

::: {.panel}

### Paperpile {.panel-name}

asdf

:::

:::::